Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-27 paper.tex

In previous versions of \msprime, silent mutations were disallowed,
and we could have removed them from the output entirely.
However, we have chosen to leave them in, so that for instance
simulating with the \texttt{HKY} mutation model will result in silent mutations
if not all equilibrium frequencies are the same.
Although the presence of silent mutations may at first be surprising
(although they don't affect calculations of statistics in any way),
there is a good reason to leave them in:
to allow layering of different mutation models.
Suppose that we wanted to model the mutation process as a mixture of more than one model,
e.g., Jukes-Cantor mutations at rate $\mu_1$, and
HKY mutations occur at rate $\mu_2$.
Layering multiple calls to \texttt{sim_mutations} is allowed,
so we could first apply mutations with the \texttt{JC69} model at rate $\mu_1$
and then add more with the \texttt{HKY} model at rate $\mu_2$.
However, there is a small statistical problem:
suppose that after applying Jukes-Cantor mutations we have an $A \to C$ mutation,
but then the HKY mutations inserts another mutation in the middle,
resulting in $A \to C \to C$.
If neither mutation model allows silent transitions,
then this is clearly not correct,
i.e., it is not equivalent to a model that simultaneously applies the two models.
(The impact is small, however, as it only affects sites with more than one mutation.)
The solution is to make the Jukes-Cantor model \emph{state-independent}
(also called ``parent-independent''),
by placing mutations at rate $4/3 \mu_1$ and then choosing the derived state for each mutation
\emph{independently} of the parent (so that 1/4 of mutations will be silent).
If so -- and, more generally, if the first mutational process put down is state-independent --
then the result of sequentially applying the two mutation models
is equivalent to the simultaneous model.
To facilitate this, many mutation models have a \texttt{state\_independent} option
that increases the number of silent mutations
and makes the model closer to state-independent.

@@

\begin{eqnarray*}
	p'=\left\{ \begin{array}{ll}
	    p + \mu(p)\delta t +\sqrt{p(1-p)\delta t} & \quad \mbox{with probability $1/2$}\\
	    p + \mu(p)\delta t -\sqrt{p(1-p)\delta t} & \quad \mbox{with probability $1/2$}\\
	\end{array} \right.
\end{eqnarray*}
where $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
assumes genic selection (i.e., that the dominance coefficient $h=0.5$. Further
$\mu(p)$ is
\begin{eqnarray*}
	\mu(p)=\frac{\alpha p (1 - p)}{tanh(\alpha(1-p))}
\end{eqnarray*}

>>>>

In previous versions of \msprime, silent mutations were disallowed,
and we could have removed them from the output entirely.
However, we have chosen to leave them in, so that for instance
simulating with the \texttt{HKY} mutation model will result in silent mutations
if not all equilibrium frequencies are the same.
Although the presence of silent mutations may at first be surprising
(although they don't affect calculations of statistics in any way),
there is a good reason to leave them in:
to allow layering of different mutation models.
Suppose that we wanted to model the mutation process as a mixture of more than one model,
e.g., Jukes-Cantor mutations at rate $\mu_1$, and
HKY mutations occur at rate $\mu_2$.
Layering multiple calls to \texttt{sim\_mutations} is allowed,
so we could first apply mutations with the \texttt{JC69} model at rate $\mu_1$
and then add more with the \texttt{HKY} model at rate $\mu_2$.
However, there is a small statistical problem:
suppose that after applying Jukes-Cantor mutations we have an $A \to C$ mutation,
but then the HKY mutations inserts another mutation in the middle,
resulting in $A \to C \to C$.
If neither mutation model allows silent transitions,
then this is clearly not correct,
i.e., it is not equivalent to a model that simultaneously applies the two models.
(The impact is small, however, as it only affects sites with more than one mutation.)
The solution is to make the Jukes-Cantor model \emph{state-independent}
(also called ``parent-independent''),
by placing mutations at rate $4/3 \mu_1$ and then choosing the derived state for each mutation
\emph{independently} of the parent (so that 1/4 of mutations will be silent).
If so -- and, more generally, if the first mutational process put down is state-independent --
then the result of sequentially applying the two mutation models
is equivalent to the simultaneous model.
To facilitate this, many mutation models have a \texttt{state\_independent} option
that increases the number of silent mutations
and makes the model closer to state-independent.

@@

\begin{eqnarray*}
	p'=\left\{ \begin{array}{ll}
	    p + \mu(p)\delta t +\sqrt{p(1-p)\delta t} & \quad \mbox{with probability $1/2$}\\
	    p + \mu(p)\delta t -\sqrt{p(1-p)\delta t} & \quad \mbox{with probability $1/2$}\\
	\end{array} \right.
\end{eqnarray*}
where $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
assumes genic selection (i.e., that the dominance coefficient $h=0.5$). Further
$\mu(p)$ is
\begin{eqnarray*}
	\mu(p)=\frac{\alpha p (1 - p)}{tanh(\alpha(1-p))}
\end{eqnarray*}
